---
title: "Water Fectching and Child Development"
subtitle: "Sub-Saharan Africa, 2000-2015"
author: "Austin Hart"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  ioslides_presentation:
    widescreen: true
    css: iocustom.css
---

```{r setup, include=FALSE}
  options(htmltools.dir.version = FALSE)
  knitr::opts_chunk$set(
    message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
    fig.align = 'center'
  )
```

```{r results='hide'}
  library(tidyverse)
  library(scales)

# data: DHS
  df = haven::read_dta('ssa water.dta') %>%
    mutate(across(where(haven::is.labelled), ~as_factor(.)))
  
# plot theme
  mytheme = theme_bw() + theme(
    plot.title = element_text(hjust = 0, size = 16, face = 'bold'),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(margin = unit(c(3, 0, 0, 0), "mm"), size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(margin = unit(c(0, 3, 0, 0), "mm"), size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = 'gray80',linetype = 'solid',size = 0.35),
    panel.grid.major.x = element_blank(),
    axis.ticks = element_blank(),
    panel.spacing = unit(2, "lines"),
    plot.margin=unit(c(.2,.2,.2,.2),"cm"),
    panel.border = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = 'bold')
  )  

```


# Methodology


## Project data

### USAID Demographic and Health Surveys (DHS)

- Data from all DHS Modules, 2000-2015
- Household surveys in `r nrow(count(df, country))` SSA countries
- Records for `r format(nrow(df), big.mark = ',')` children, aged 5-14
- Block randomized sampling within country
    * Geographic clusters chosen at random
    * Random sampling among proximate households


# Water fetching times

## Fetching times higher among rural households

> Half of urban HHs walk 3 minutes or less to water. 1 in 3 rural HHs more than 30 minutes away.

```{r wfurban}

  df %>%
    filter(watertime < 180) %>%
    ggplot(aes(x = 1+watertime, color = rural, fill = rural)) + # 1+ for log10 transform
    geom_density(aes(y = stat(0.08 * density)), # change axis to percent
        bw = 0.08, position = 'dodge', alpha = .25, size = 1) +
    scale_x_log10(breaks = c(0,5,10,30,60,90),labels = c(0,5,10,30,60,90)) +
    scale_fill_viridis_d(direction = -1, aesthetics = c('fill','color')) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      x = 'Walk time to water (minutes)',
      y = NULL,
      color = NULL,
      fill = NULL,
      title = "Walk time, rural vs urban households"
    ) + mytheme
```

## Wealth buys proximity to water

> Distance to water significantly lower among wealthiest quintiles.

```{r wealth}

  df %>%
    #start with group medians
    filter(!is.na(wealth)) %>%
    group_by(wealth) %>%
    summarise(
      MedianWalk = median(watertime, na.rm = TRUE)
    ) %>%
  #now plot  
    ggplot(aes(x = wealth, y = MedianWalk)) +
      geom_hline(yintercept = 0, color = 'gray60') +
      geom_segment(aes(xend = wealth, yend = 0), size = .85) +
      geom_point(shape = 22, color = 'black', fill = 'dodgerblue2', size = 12) +
      geom_text(aes(label = MedianWalk),
                color = 'white', size = 4) +
  # beautify
      scale_y_continuous(breaks = seq(0,20,10)) +
      coord_cartesian(ylim = c(0,25)) +
      labs(
        title = 'Median walk time by income quintile',
        x = NULL,
        y = 'Walk to water source, mins'
      ) +
      mytheme +
      theme(axis.text.y = element_blank())
```


# Water fetching and childhood education

## School attendance invariant to walk time

> Among **urban** HHs, school attendance is slightly higher for boys, but water fetching time is unrelated to attendance for boys or girls.


```{r attend}

  df %>%
    filter(watertime < 120 & rural == 'urban') %>%
  # make the plot
    ggplot(aes(y = as.numeric(attend) - 1, x = 1 + watertime, #+1 to enable log scale
               color = female, linetype = female)) + 
    geom_smooth(method = 'lm',se = F) +
  # fix axes
    scale_y_continuous(
      limits = c(0,1), breaks = c(0,.5,1),
      labels = scales::percent
    ) +
    scale_x_log10(
      breaks = c(1,5,10,30,60,120),
      labels = c(0,5,10,30,60,120)
    ) +
  # beautify
    scale_color_manual( values = c('purple','black') ) +
    labs(
      title = 'Attendance over walk time by gender, urban homes',
      x = 'Walk time (minutes)',
      y = 'Attending school',
      color = NULL,
      linetype = NULL
    ) +
    mytheme +
    theme(
      panel.grid.major.x = element_line(color = 'gray80'),
      legend.position = 'top'
    )
```

# RMarkdown Reminders

## Setup

### Markdown documents need:

- YAML header
- Setup code chunk
- Body of text

## YAML header

- Specifies title, author, date, etc
- Specifies the output format
- **Spacing/indentation are critical**

```{r yaml, eval=F, echo=T}
---
title: "Water Fectching and Child Development"
subtitle: "Sub-Saharan Africa, 2000-2015"
author: "Austin Hart"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  ioslides_presentation:
    widescreen: true
    css: iostyle.css
---
```

## Setup chunk

- Defines behavior for ALL other code chunks
- **Must** come first
- Label explicitly as `setup`

````
```{r setup, include=FALSE}`r ''`
  options(htmltools.dir.version = FALSE)
  knitr::opts_chunk$set(
    message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE, 
    fig.align='center'
  )
```
````

## Body of doc/deck

Everything else is a combo of text and `R` code chunks.

Remember:

- Code chunks require *unique* names
- `Knit` often to diagnose errors
- Explore different outputs, esp. `Beamer` (requires `tinytex`)
